\documentclass[12pt,a4paper]{article}

% --- Encoding & fonts (XeLaTeX) ---
\usepackage{fontspec}
\setmainfont{Times New Roman}
\setmonofont{Courier New}

% --- Language ---
\usepackage{polyglossia}
\setdefaultlanguage{english}

% --- Math ---
\usepackage{amsmath,amssymb,amsthm}
\usepackage{unicode-math}
\usepackage{booktabs}
\usepackage{array}
\usepackage{tabularx}
\usepackage{graphicx}
\usepackage{caption}
\captionsetup{labelfont=bf, textfont=it}
\usepackage{longtable}

% --- Page layout ---
\usepackage[a4paper, margin=2.5cm, bottom=3cm]{geometry}
\usepackage{setspace}
\onehalfspacing
\raggedbottom

% --- Typography ---
\usepackage{microtype}
\usepackage[stable]{footmisc}
\frenchspacing

% --- Running header / footer ---
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small A Structural Microsimulation of a Neighborhood Grocery Store}
\fancyhead[R]{\small Chinh Mai}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

% --- Hyperlinks ---
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

% --- Lists ---
\usepackage{enumitem}
\setlist{leftmargin=*, topsep=6pt, itemsep=4pt, parsep=0pt}

% --- Diagrams ---
\usepackage{tikz}
\usetikzlibrary{positioning, arrows.meta, shapes.geometric, fit, backgrounds}

\title{\textbf{A Structural Microsimulation of a Neighborhood Grocery Store:\\
Theoretical Foundations, Model Architecture, and Design Rationale}}
\author{Chinh Mai}
\date{July 2026}

\begin{document}

\maketitle

\begin{abstract}
This paper documents the theoretical foundations of a synthetic business data
generator built to produce a three year history of a small neighborhood
grocery store together with a fully known generating process. Rather than
sampling numbers directly, the generator constructs a world: a street with
locations, households with preferences and budgets, an owner with capital and
beliefs, a calendar of seasons and shocks, and a daily market in which the two
sides interact hour by hour. Every distributional choice in that world is
defended by the same three question discipline, namely what the variable's
support permits, what generative story produces it, and which distribution
adds the least unwarranted structure beyond what is claimed. This paper
consolidates that reasoning across the five construction phases, from the
opening decision solved as a mixed integer program through the three year
panel of customer turnover and capital accumulation, and explains why each
structural choice was made rather than merely what it computes. The result is
intended as a companion reference for readers who want to understand the
model itself, independent of the software that implements it.
\end{abstract}

\clearpage

\section{Introduction}

Every real business dataset is the residue of a world. Behind a spreadsheet of
receipts and invoices sit people with habits and budgets, a street with rents
and foot traffic, and an owner with limited capital and imperfect information,
all colliding day after day until the paperwork accumulates. A dataset built
to teach or test data analysis honestly should therefore not be built by
drawing numbers from convenient distributions and calling the result a
business. It should be built the other way around: construct the world first,
with its people, its constraints, and its causal structure, and let the
paperwork fall out of the world's own operation.

This paper documents the theoretical reasoning behind such a construction. The
subject is a simulated small grocery store, run by a single owner over three
years, generating the full range of records a real shop produces: till
receipts, supplier invoices, shelf price histories, promotional campaigns,
monthly ledgers, and a tax filing. What distinguishes the dataset from an
arbitrary spreadsheet of plausible numbers is that every number has a
traceable cause. Prices respond to costs, costs respond to shocks, shoppers
respond to prices according to individually varying tastes, and the owner
responds to what he observes, which is never the same as what is true. The
gap between what the owner believes, what actually happens, and what a
perfectly informed operator could have achieved is not injected noise; it is
the mechanical consequence of a chain of modeling decisions, each of which
this paper defends on its own terms.

The construction proceeds in five phases, summarized in Figure~\ref{fig:pipeline}
and detailed in the sections that follow. The first phase builds the world as
it exists the moment before the store opens: a street of candidate locations,
a population of households with preferences, and an owner who must decide
where to open, what to stock, and how much capital to commit, all before a
single customer has walked through a door. The second phase writes the
script of the year that follows, meaning every exogenous influence on the
world, such as the weather, the calendar, and macroeconomic shocks, decided in
advance and independent of anything the simulated agents do. The third phase
performs that script: customers and the owner interact hour by hour, and
every transaction, stockout, and promotional decision is recorded. The fourth
phase turns the single performed year into a laboratory, allowing the same
world to be replayed under edited conditions so that the causal effect of a
tax reform or a competitive shock can be measured exactly rather than
estimated. The fifth phase extends the horizon to three years, adding customer
turnover, capital accumulation, and a sequence of scripted events that a real
small business might plausibly encounter.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance = 10mm and 8mm,
    box/.style = {rectangle, draw, rounded corners, align=center, minimum width=2.6cm, minimum height=1.15cm, font=\small},
    lbl/.style = {font=\scriptsize, align=center, text width=2.6cm},
    arrow/.style = {-{Latex[length=2mm]}, thick}
]
    \node[box] (p1) {Phase 1\\[2pt]The world\\at $t=0$};
    \node[box, right=of p1] (p2) {Phase 2\\[2pt]The script\\of the year};
    \node[box, right=of p2] (p3) {Phase 3\\[2pt]The daily\\market loop};
    \node[box, right=of p3] (p4) {Phase 4\\[2pt]The policy\\laboratory};
    \node[box, below=of p4] (p5) {Phase 5\\[2pt]Three years:\\panel and capital};

    \draw[arrow] (p1) -- (p2);
    \draw[arrow] (p2) -- (p3);
    \draw[arrow] (p3) -- (p4);
    \draw[arrow] (p4) -- (p5);

    \node[lbl, below=2mm of p1] {locations, customers,\\the opening decision};
    \node[lbl, below=2mm of p2] {calendar, weather,\\macro shocks};
    \node[lbl, below=2mm of p3] {transactions,\\hidden demand};
    \node[lbl, above=2mm of p4] {edited scripts,\\taxes};
    \node[lbl, below=2mm of p5] {churn, retained\\earnings, growth};
\end{tikzpicture}
\caption{The five phase construction. Each phase consumes the outputs of the ones before it and adds one new kind of structure to the world.}
\label{fig:pipeline}
\end{figure}

The remainder of this paper is organized around that same five phase
structure. Section~\ref{sec:literature} positions this construction against
the other ways synthetic and case study business data are currently produced,
and discusses what the resulting dataset makes possible, both analytically
and pedagogically. Section~\ref{sec:principles} states the design principles
and notational conventions that recur throughout the model. Sections
\ref{sec:phase1} through \ref{sec:phase5} treat each phase in turn, explaining
not only how each mechanism is built but why it was built that way rather
than some simpler or more elaborate alternative. Section~\ref{sec:accounting}
describes the accounting layer that ties every euro and every unit of stock
to a specific artifact, including the deliberate imperfections that make the
raw records disagree until they are cleaned. Section~\ref{sec:discussion}
reflects on the architecture as a whole, and Section~\ref{sec:conclusion}
closes the paper.

\section{Related Approaches and the Case for Structural Synthesis}
\label{sec:literature}

Synthetic business data is not a new idea, and it is worth being precise
about where this construction sits relative to the approaches already in
routine use, rather than presenting it as though it emerged in a vacuum.
Four traditions dominate the current landscape, each built to optimize for
something other than what this project optimizes for.

\subsection{Rule based generators}

The simplest and most widely used tools, such as Faker and Mockaroo, fill a
schema with plausible looking field values: names, addresses, dates,
amounts drawn from simple parametric distributions attached independently to
each column. These tools are fast, require no seed data, and are entirely
adequate for their actual purpose, which is populating a database or testing
a software pipeline against realistically shaped inputs. They are not built
to support analysis practice, because each column is generated
independently of every other: there is no mechanism linking a customer's
income to their basket size, no mechanism linking weather to foot traffic,
and no sense in which the resulting numbers were caused by anything. A
student who regresses one generated column on another is not discovering a
relationship, because none was ever built in; every correlation
Section~\ref{sec:phase1} through Section~\ref{sec:phase5} of this paper
painstakingly justifies is, in a rule based generator, simply absent by
construction.

\subsection{Statistical and generative model synthesizers}

A more sophisticated family of tools learns to imitate the statistical
shape of a real dataset rather than sampling columns independently.
Conditional generative adversarial networks such as CTGAN, variational
autoencoders, and copula based tools bundled in synthesis libraries such as
the Synthetic Data Vault (\hyperlink{ref:patki2016}{Patki, Wedge, and
Veeramachaneni, 2016}; \hyperlink{ref:xu2019}{Xu, Skoularidou,
Cuesta-Infante, and Veeramachaneni, 2019}) are trained on a real
seed dataset and learn to reproduce its marginal distributions and pairwise
correlations closely enough to substitute for it in downstream use, often
motivated by a differential privacy guarantee
(\hyperlink{ref:dwork2006}{Dwork, 2006}) that lets an
organization share a synthetic look-alike without exposing real customer
records. This is a genuine and valuable technology for its actual purpose,
which is privacy preserving data sharing. It is a poor fit for teaching data
analysis for three specific reasons. First, it requires an existing real
dataset to imitate, so it cannot manufacture a genuinely new case; every
synthetic version is a shadow of something that already happened elsewhere.
Second, the generator itself is a black box: a neural network or a fitted
copula reproduces a distribution's shape without asserting any mechanism
that produced it, so there is no known ground truth against which a
student's causal claim, elasticity estimate, or forecast can be graded,
only the observed data the student was already given. Third, and most
importantly for this project's purposes, a distribution matching model has
no way to answer a counterfactual question, because it was never given a
reason for any number, only a shape to reproduce; asking such a model what
revenue would have been without a tax change is not merely hard, it is
category confusion, since the model has no representation of a tax change
to remove in the first place.

\subsection{The business case method}

Business schools have long taught with hand authored cases: a narrative
description of a company's situation accompanied by a set of exhibits,
built by a case writer around a specific pedagogical point. This tradition,
associated most closely with Harvard Business School's case method, remains
the gold standard for teaching judgment under ambiguity, and the case
narrative structure adopted for this project's own teaching materials
borrows directly from it. Its limitation is not narrative quality but
process: each case is bespoke, manually tuned by its author to make the
numbers come out right, produced once, and then reused unchanged for years.
There is no underlying generative process to interrogate, no way to produce
a counterfactual twin of the same company under a different tax regime, and
no way to grade a student's causal claim against a documented mechanism,
because the case writer's spreadsheet is the only ground truth that exists
and it was authored to support a conclusion rather than to allow one to be
discovered.

\subsection{Structural microsimulation}

The methodological tradition closest to this project's own is structural
microsimulation, founded by \hyperlink{ref:orcutt1957}{Orcutt's (1957)}
proposal to model an economy as
a population of individual units following explicit behavioral rules rather
than as aggregate equations, and carried forward today in tax and benefit
microsimulation models such as EUROMOD and in agent based market and
transport simulations across applied economics. These models share this
project's central commitment: agents obey genuine, documented behavioral
rules, so a policy change can be replayed through an unchanged population
and the difference is a true causal effect rather than an estimated one,
exactly the property Section~\ref{sec:phase4} builds into this model's own
policy laboratory. Where such models typically diverge from this project is
in what they are built to produce. Their output is usually an aggregated
statistic destined for a policy briefing, such as a national poverty rate
or a revenue projection, rather than a transaction level business record;
and because their audience is professional economists rather than students
learning to clean and reconcile data, they rarely simulate the imperfect
paperwork of Section~\ref{sec:phase3}, since a policy model has no reason to
misfile an invoice. Structural microsimulation gets the causal architecture
right; it has simply never been aimed at the specific problem of teaching
an individual analyst to reconstruct a single business's truth from its own
imperfect records.

\subsection{What this construction combines}

No single tradition surveyed above combines all four of the properties this
project requires simultaneously. It requires, first, a fully documented
structural generating process rather than a black box, so that every
pattern in the data has a named cause, in the spirit of structural
microsimulation rather than distributional synthesis. It requires, second,
transaction level business artifacts, till receipts and supplier invoices
rather than aggregated statistics, carrying deliberately realistic and
separately documented paperwork imperfections, in the spirit of a genuine
business record rather than a clean simulation output or a case writer's
tidy exhibit. It requires, third, exact counterfactual replay at the level
of individual decisions, made possible only by the keyed randomness
discipline of Section~\ref{sec:principles}, which neither a hand authored
case nor a distributional synthesizer can offer at all. It requires,
fourth, a built in scorecard for the analysis itself: the three profit
figures of Section~\ref{sec:phase3} turn the abstract idea of analytical
value into a specific number, the gap between what the owner believed, what
he actually earned, and what a perfectly informed operator could have
earned, which is not a feature any of the four traditions above were ever
designed to provide, because none of them stage a protagonist who is
wrong in a way that can be measured.

\subsection{The potential of the resulting data}

A dataset built this way supports a genuine progression of analytical
questions rather than a single fixed exercise, because the same underlying
mechanisms answer descriptive, diagnostic, predictive, prescriptive, and
counterfactual questions alike, each gradeable against the hidden truth the
generator itself recorded. At the descriptive layer, a class can ask where
the money comes from and goes, when people shop, and what a basket looks
like, all directly readable from the ledger and the receipts. At the
diagnostic layer, the same data supports genuine causal questions with
known answers: whether weather actually moves revenue once the calendar is
controlled for, exactly the mechanism of Section~\ref{sec:phase2}; how much
of a wholesale cost shock reaches the shelf and how quickly, measurable
against the menu cost rule of the same section; and whether a price increase
loses customers outright or merely redirects them to a close substitute,
answerable against the discrete choice structure of
Section~\ref{sec:phase1}. At the predictive layer, a forecasting exercise
can be benchmarked not only against a held out quarter but against the
owner's own documented forecasting rule, exposing exactly where and why a
naive trailing average underperforms, per Section~\ref{sec:phase3}. At the
prescriptive layer, a class can redesign the ordering policy or the markdown
trigger and price the redesign in real euros against the oracle figure of
Section~\ref{sec:phase3}, rather than merely arguing that a better policy
would probably help. At the counterfactual layer, the scenario twins of
Section~\ref{sec:phase4} and the three year arc of Section~\ref{sec:phase5}
let a class ask what a tax cut actually did to prices, what a discounting
competitor actually cost, and whether an expansion decision paid for itself,
each answerable as an exact difference between two runs of the same world
rather than an inference from a single history. No other synthetic data
source surveyed above supports this particular ladder from description
through causal counterfactual with a documented answer waiting at every
rung.

\subsection{Value for teaching scientific reasoning, not only data analysis}

The deeper claim behind this construction is that its principal teaching
value is not technical fluency with a particular software tool but a
disciplined scientific habit of mind, and that a dataset with a fully known
generating process is uniquely suited to teaching that habit because it can
tell a student, definitively, when they are right. An analyst working this
data is repeatedly placed in a position where a plausible looking pattern is
a trap rather than a finding: the flat category of Section~\ref{sec:phase2}
whose apparent seasonality is a budget mediation effect rather than a
genuine preference, the markdown campaign of Section~\ref{sec:phase3} whose
naive before and after comparison is contaminated by the very selection rule
that triggered it, and the energy crisis of Section~\ref{sec:phase2} whose
use as a price instrument is subtly invalid because it also moves household
budgets. Each of these traps has a documented, correct resolution, which
means an instructor can teach the general scientific skill of interrogating
one's own inference, not by asserting that skepticism is a virtue in the
abstract, but by handing a student a specific claim, a specific method for
checking it, and a specific ground truth that reveals whether the checking
actually worked. The same property extends beyond individual traps to the
whole discipline of hypothesis formation: because the model is built from
named mechanisms defended by the three question discipline of
Section~\ref{sec:principles} rather than fitted after the fact to match a
target, a student can be asked to propose a mechanism before seeing whether
the data supports it, and then be told, precisely, whether they were right,
which is the same discipline a working analyst must apply to a real
business where no such answer key will ever be handed over. This is,
ultimately, the justification for building the world first and letting the
paperwork fall out of its own operation, as this paper's introduction
proposed: only a dataset built that way can teach an analyst not merely how
to compute an estimate, but how to know whether an estimate deserves to be
believed.

\section{Design Principles and Notation}
\label{sec:principles}

\subsection{The three question discipline}

Every stochastic element of the model, from a customer's weekly grocery
budget to the day a storm arrives, was chosen by answering the same three
questions, always in the same order.

\begin{enumerate}
\item \textbf{Support.} What values can the variable actually take? A
      probability must live on $(0,1)$, a count must be a nonnegative
      integer, a share of a composite basket must live on a simplex, and a
      cost must be strictly positive. The distribution chosen must respect
      this support without needing an artificial truncation.
\item \textbf{Generative story.} What process actually produces the
      variable? A quantity that arises from many small effects adding
      together behaves differently from one that arises from many small
      factors multiplying together, and both differ again from a quantity
      that arises from a fixed number of independent yes or no trials at a
      common rate.
\item \textbf{Maximum entropy.} Once the support and the generative story
      have narrowed the field, which distribution within that field adds the
      least unwarranted structure beyond the facts one is actually willing to
      claim? This is the maximum entropy principle in its practical form: if
      the only facts asserted about a variable are its mean and its variance,
      the distribution that respects exactly those two facts and assumes
      nothing else is the least presumptuous choice available.
\end{enumerate}

This discipline explains a recurring pattern in the model. Wherever a
quantity is the product of many small positive multiplicative factors, such
as a household's grocery budget (itself a product of household size, income
level, and consumption habits), the natural distribution is the LogNormal,
because summing the logarithms of many small factors is exactly the setting
in which the central limit theorem produces a Normal distribution, and
exponentiating a Normal produces a LogNormal that is guaranteed positive.
Wherever a quantity is a probability, meaning it must live strictly between
zero and one, the Beta distribution is the natural family, because it is the
maximum entropy distribution on that support once a shape is fixed, and its
two parameters give direct control over where the mass concentrates. Wherever
a quantity counts independent rare events at a common rate, the Poisson
distribution follows, because it is the maximum entropy distribution for
counts given only a rate. These are not arbitrary choices repeated by habit;
each one answers the same three questions afresh in its own context, and each
answer is defended at the point where the variable first appears.

\subsection{Notation}

The model follows a McElreath style notation throughout, meaning every
distribution is written with its natural parameters, and any scale parameter
is a standard deviation rather than a variance, so that a Normal distribution
is always written $\text{Normal}(\mu, \sigma)$ and never
$\text{Normal}(\mu, \sigma^2)$. Three kinds of assignment recur.

\begin{enumerate}
\item The symbol $\sim$ denotes a stochastic draw, as in $X \sim
      \text{Beta}(2,2)$, meaning $X$ is a random realization from that
      distribution.
\item The symbol $=$ denotes a deterministic definition, as in $D_{cl} =
      \text{Households}_l \cdot \mu_c$, meaning the left hand side is
      computed exactly from the right hand side with no additional
      randomness.
\item The symbol $\overset{\leftarrow}{=}$ is used specifically where the
      direction of causation is itself the point being made, most often in
      the cost equations, to mark that the right hand side causes the left
      hand side rather than merely equaling it algebraically.
\end{enumerate}

A hat, written $\hat{\cdot}$, always marks a belief held by the owner, as
distinct from the corresponding true value that the owner cannot observe.
Indices recur across the whole model: locations are indexed by $l$,
categories of merchandise by $c$ (there are twelve, matching a real product
catalog), individual products or stock keeping units by $s$, customers by
$i$, and time by $t$, measured in days across a horizon that eventually spans
three years, with weeks indexed by $w$, months by $m$, and individual shock
events by $k$. A small number of letters, such as $\alpha$ and $m$, are
reused across different phases for unrelated quantities, exactly as they are
reused in the underlying source material; every such reuse is flagged in
prose at the point where it recurs, and the table below lists each meaning
separately under the section where it first appears.

Every symbol used in a displayed formula anywhere in this paper is defined
in prose at the point it first appears. Because the paper is long and a
symbol's definition can be several pages before a later formula that reuses
it, Table~\ref{tab:notation} collects every symbol in one place, organized
by the section in which it is first introduced, purely as a reference for
checking a formula against its plain language meaning without having to
search back through the text.

{\footnotesize
\begin{longtable}{@{}p{2.3cm}p{9.6cm}p{2.3cm}@{}}
\caption{Notation used throughout the paper, grouped by the section in which each symbol first appears.}
\label{tab:notation} \\
\toprule
\textbf{Symbol} & \textbf{Meaning} & \textbf{First used} \\
\midrule
\endfirsthead
\toprule
\textbf{Symbol} & \textbf{Meaning} & \textbf{First used} \\
\midrule
\endhead
\bottomrule
\endfoot
\multicolumn{3}{@{}l}{\textit{Indices and meta-notation}} \\
$l$, $c$, $s$, $i$, $t$ & location, category, product/SKU, customer, and day indices & \S\ref{sec:principles} \\
$w$, $m$, $k$ & week, month, and shock-event indices & \S\ref{sec:phase2}--\ref{sec:phase5} \\
$\hat{\cdot}$ & a belief held by the owner, as opposed to the true value & \S\ref{sec:principles} \\
$\sim$, $=$, $\overset{\leftarrow}{=}$ & stochastic draw; deterministic definition; causal definition & \S\ref{sec:principles} \\
\addlinespace
\multicolumn{3}{@{}l}{\textit{Phase One --- locations and the opening decision}} \\
$Q_l$ & latent quality score of location $l$, on $[0,1]$ & \S\ref{sec:phase1} \\
Households$_l$, Rent$_l$, SetupCost$_l$ & a location's catchment size, monthly rent, and fit-out cost & \S\ref{sec:phase1} \\
$N_l$ & pool of households at location $l$ who adopt the store & \S\ref{sec:phase1} \\
$B_i$ & customer $i$'s baseline weekly grocery budget & \S\ref{sec:phase1} \\
$d_i$, $\pi_i$, $p_i$ & primary shopping day; probability of shopping on it; probability of an extra midweek visit & \S\ref{sec:phase1} \\
$w_i$, $\alpha$ & basket composition across categories; its Dirichlet concentration vector & \S\ref{sec:phase1} \\
$\beta_i$ & customer $i$'s price sensitivity & \S\ref{sec:phase1} \\
$b_i$, $z_i$, $\nu_i$ & brand affinity; standardized log budget; independent brand-affinity noise & \S\ref{sec:phase1} \\
$T_i$, PaysByCard$_{iv}$ & card-preferring type; whether visit $v$ is paid by card & \S\ref{sec:phase1} \\
$U_{is}$ & customer $i$'s utility for product $s$ & \S\ref{sec:phase1} \\
$\theta_{ic}$ & customer $i$'s current need for category $c$ & \S\ref{sec:phase1} \\
$a_s$, $\gamma$ & product $s$'s unmodeled appeal intercept; weight on brand fit & \S\ref{sec:phase1} \\
BrandLevel$_s$ & product $s$'s position on the brand spectrum & \S\ref{sec:phase1} \\
$D_{cl}$, $\mu_c$ & true demand potential for category $c$ at location $l$; mean monthly per-household consumption of category $c$ & \S\ref{sec:phase1} \\
$\hat{D}_{cl}$, $\delta_c$ & owner's believed demand; his log-scale belief error & \S\ref{sec:phase1} \\
$m_c$ & category $c$'s fixed retail markup & \S\ref{sec:phase1} \\
$\eta$, $\rho$ & safety-stock fraction ($0.30$); cap on one product's share of category demand ($0.05$) & \S\ref{sec:phase1} \\
TotalCost$_0$, FixedCost$_0$, VariableCost$_0$ & opening cost components, decomposed causally & \S\ref{sec:phase1} \\
$y_l$, $x_s$ & binary location choice; binary listing indicator & \S\ref{sec:phase1} \\
$q_s$, $u_s$, $E^h_0$ & units purchased at opening; believed units sold; extra staff hired & \S\ref{sec:phase1} \\
\addlinespace
\multicolumn{3}{@{}l}{\textit{Phase Two --- calendar, weather, and cost shocks}} \\
$T_t$, $\mu_T(t)$, $\varepsilon_t$, $\phi$ & day $t$'s temperature; its seasonal cycle; its persistent anomaly; the anomaly's AR(1) persistence ($0.7$) & \S\ref{sec:phase2} \\
$W_t$, $p_{01}$, $p_{11}$ & whether it rains on day $t$; dry-to-wet and wet-to-wet transition probabilities & \S\ref{sec:phase2} \\
$R_t$ & rainfall amount on a wet day & \S\ref{sec:phase2} \\
$M_{ct}$ & demand modifier for category $c$ on day $t$ & \S\ref{sec:phase2} \\
$a_c$, $\kappa_c$, $h_c$ & category $c$'s seasonal amplitude; weather-anomaly loading; pre-holiday loading & \S\ref{sec:phase2} \\
$z_t$, $H_t$ & standardized temperature anomaly; pre-holiday flag & \S\ref{sec:phase2} \\
$\psi_{pt}$, $a_p$, $\kappa_p$ & product-type $p$'s own seasonal tilt, amplitude, and anomaly loading & \S\ref{sec:phase2} \\
$\Lambda_t$, $\phi^\Lambda$ & daily foot-traffic modifier; its AR(1) persistence ($0.85$) & \S\ref{sec:phase2} \\
$B_{iw}$, $\xi_{iw}$ & customer $i$'s week-$w$ budget; its small weekly wobble & \S\ref{sec:phase2} \\
$s_{iw}$, $g_{iw}$ & tight-spell multiplier; splurge multiplier, both on $B_{iw}$ & \S\ref{sec:phase2} \\
$\bar{g}_t$ (or $\bar{g}_w$) & the scripted energy crisis's own trajectory over time & \S\ref{sec:phase2} \\
$\text{infl}(t)$ & the continuous annual inflation drift & \S\ref{sec:phase2} \\
$K$, $\tau_k$ & number of shock events in the year; arrival day of event $k$ & \S\ref{sec:phase2} \\
$\zeta_k$, $\rho_k$ & peak magnitude of shock $k$; its exponential decay time constant & \S\ref{sec:phase2} \\
$g_k(t)$, $A_k$ & shock $k$'s effect on day $t$; the set of categories it touches & \S\ref{sec:phase2} \\
CostMult$_{ct}$ & wholesale cost multiplier for category $c$ on day $t$ & \S\ref{sec:phase2} \\
$\overline{\text{Cost}}_{s,t}$, $\alpha$ & owner's smoothed invoice-cost trend for product $s$; its EWMA weight ($0.35$, unrelated to the basket $\alpha$ above) & \S\ref{sec:phase2} \\
\addlinespace
\multicolumn{3}{@{}l}{\textit{Phase Three --- the daily market and the recording layer}} \\
$P_{ict}$, $r_{ict}$, $r_{ic}$ & customer $i$'s pantry stock of category $c$; its daily drain rate; the drain rate's undistorted base level & \S\ref{sec:phase3} \\
$\theta_{ict}$, $P^{\text{tgt}}_{ic}$, $\tau$ & need intensity from pantry shortfall; target pantry level; target days of cover ($10$) & \S\ref{sec:phase3} \\
$\alpha$ (need shape) & shape constant of the pantry-shortfall need function, unrelated to the two earlier uses of $\alpha$ & \S\ref{sec:phase3} \\
$\kappa$ (purchase) & random fraction of the shortfall actually purchased, Uniform$(0.7,1.0)$ & \S\ref{sec:phase3} \\
$\hat{D}^{\text{op}}_{cw}$ & owner's trailing four-week sales forecast for category $c$ & \S\ref{sec:phase3} \\
Target$_{sw}$, $R$, $L$ & order-up-to target for product $s$; review period ($7$ d); lead time ($2$ d) & \S\ref{sec:phase3} \\
$\tilde{\sigma}_{sw}$ & product $s$'s smoothed share of its category's recent sales & \S\ref{sec:phase3} \\
Spoiled$_{st}$, $\lambda_c$, $\lambda^d_c$ & units of product $s$ spoiled overnight; category's weekly base spoilage rate; its converted nightly rate & \S\ref{sec:phase3} \\
$f_{ct}$, $\kappa^T_c$, $\kappa^E_c$, $\bar{T}$ & spoilage factor; heat-term loading; cold-chain loading; annual mean temperature & \S\ref{sec:phase3} \\
$D_{st}$, $B_{st}$, $S_{st}$, $A_{sc}$ & book-minus-true discrepancy; book stock; true stock; monthly count correction & \S\ref{sec:phase3} \\
$\pi_{\text{believed}}$, $\pi_{\text{realized}}$, $\pi_{\text{oracle}}$ & owner's believed profit; the ledger's realized profit; a perfectly informed operator's oracle profit & \S\ref{sec:phase3} \\
$\Delta_{\text{optimism}}$, $\Delta_{\text{analytics}}$ & cost of the owner's overoptimism; value of better information and decisions & \S\ref{sec:phase3} \\
\addlinespace
\multicolumn{3}{@{}l}{\textit{Phase Four --- the policy laboratory and the tax layer}} \\
$r_c(t)$ & category $c$'s value added tax rate path & \S\ref{sec:phase4} \\
Tag$_{st}$, $\overline{\text{NetCost}}_{s,t}$ & product $s$'s shelf price tag; its smoothed tax-exclusive invoice cost & \S\ref{sec:phase4} \\
VAT$_t$ & net value added tax remitted in month $t$ & \S\ref{sec:phase4} \\
$r^{\text{pay}}$, $r^{\pi}$ & payroll tax rate ($0.25$); profit tax rate ($0.20$) & \S\ref{sec:phase4} \\
\addlinespace
\multicolumn{3}{@{}l}{\textit{Phase Five --- the three year panel and capital}} \\
type$_i$, $h$ & customer $i$'s persistence type (rooted or transient); transient customer's monthly departure hazard ($1/18$) & \S\ref{sec:phase5} \\
$\rho_{RE}$, $\pi_m$ & retained-earnings share ($0.5$); month $m$'s after-tax profit & \S\ref{sec:phase5} \\
RE$_m$, capex$_m$ & cumulative retained earnings through month $m$; capital spent that month & \S\ref{sec:phase5} \\
$K_{\text{exp}}$, $C_{\text{exp}}$ & retained-earnings threshold that triggers expansion; one-time fit-out cost & \S\ref{sec:phase5} \\
$m_i$, $\tilde{s}_i$ & discounter's visit-probability penalty on customer $i$ (unrelated to the markup $m_c$); customer's price-sensitivity percentile & \S\ref{sec:phase5} \\
$\chi$, $k_i$ & competitor calibration constant; rooted/transient penalty weight & \S\ref{sec:phase5} \\
\addlinespace
\multicolumn{3}{@{}l}{\textit{The accounting layer}} \\
book$(t)$, deliveries$(t)$ & the book-inventory identity's stock and delivery terms & \S\ref{sec:accounting} \\
sales$(t)$, write\_offs$(t)$ & the book-inventory identity's sales and write-off terms & \S\ref{sec:accounting} \\
ProfitBeforeTax, ProfitTax, ProfitAfterTax & the annual profit-and-tax identity's three terms & \S\ref{sec:accounting} \\
\end{longtable}
}

\subsection{Reproducibility and the discipline of keyed randomness}

The entire model is driven by a single seeded pseudorandom generator, so
that the whole three year history can be reproduced exactly from one integer.
This alone would be unremarkable were it not for a second requirement that
follows from the model's purpose as a policy laboratory (Section
\ref{sec:phase4}): the ability to replay the same world under an edited
script, such as one in which a particular shock never occurred, and to
attribute every difference between the two runs entirely to the edit.

This requirement imposes a strict discipline on how randomness is consumed.
If every random draw were taken from one sequential stream, then deleting a
single event partway through the year would shift every subsequent draw,
and the comparison between the two runs would be contaminated by reshuffled
randomness that has nothing to do with the edit itself. The model instead
keys every draw by the stable identity of the entity it concerns, such as a
specific customer on a specific day, or a specific product on a specific
night, using independently spawned child streams. A customer's Tuesday
decision is drawn from a stream indexed by that customer and that day, so
that in any counterfactual replay which does not touch that customer, the
identical draw is reproduced. This is the common random numbers technique
familiar from the simulation literature, and it is adopted from the first
line of the implementation rather than retrofitted later, because retrofitting
it after the fact would require rewriting every random draw in the model.

\section{Phase One: The World at Rest}
\label{sec:phase1}

Phase One constructs the world as it exists at the instant before the store
opens. It proceeds from both sides of the market at once. The demand side
begins with a street, populates it with households, and turns households into
customers with individually varying tastes, budgets, and shopping schedules.
The supply side begins with an entrepreneur holding a fixed sum of capital and
forming beliefs about a demand he cannot directly observe, and ends with his
first and largest decision: where to open, what to stock, how much of it to
buy, and how many people to hire, all made at once because each choice
constrains the others through the shared limits of capital and shelf space.

\subsection{The street and its locations}

Before there are customers, there is geography. The neighborhood offers a
small number of candidate premises, and they are not equal. Every observable
feature of a location, its rent, the number of households in its catchment,
its fit out cost, and its physical capacity, is modeled as flowing causally
from a single latent quality score $Q_l \in [0,1]$:

\[
\begin{aligned}
Q_l &\sim \text{Beta}(2,2) \\
\text{Households}_l &\sim \text{Normal}(250 + 400\,Q_l,\ 30) \\
\text{Rent}_l &\sim \text{Normal}(500 + 1800\,Q_l,\ 100) \\
\text{SetupCost}_l &\sim \text{Normal}(6000\,(1-Q_l),\ 400)
\end{aligned}
\]

Quality is modeled with a Beta distribution because a bounded score on
$[0,1]$ is precisely the support the Beta family occupies, and the symmetric
$\text{Beta}(2,2)$ shape encodes a defensible claim about real streets: most
premises are middling, truly excellent or truly dreadful corners are rare,
and no location is either mathematically perfect or entirely hopeless, since
the density of a $\text{Beta}(2,2)$ vanishes at both endpoints. A uniform
distribution would instead claim that a flawless location is exactly as
common as an average one, which is not a claim anyone would actually want to
make.

The Normal residuals attached to rent, household count, and setup cost each
collect the many small unmodeled influences on that quantity, such as
building size or a landlord's temperament, and the Normal distribution is the
maximum entropy choice once only a mean and a finite spread are being
claimed. The economic content of the equations is worth reading separately
from their distributional form. Household count is the causal bridge from
geography to demand, and it is the reason good locations command high rent:
rent and foot traffic share a common cause in location quality, and neither
directly causes the other. Setup cost falls as quality rises because
run-down premises with cheap rent need new wiring and refrigeration, which
plants a negative correlation between two cost components that a naive
correlation analysis would otherwise misread as a direct relationship, when
in fact both are consequences of the same underlying quality.

\subsection{Customers: budgets, schedules, and tastes}

Not every household in a location's catchment becomes a customer. The pool of
potential regulars is drawn as
$N_l \sim \text{Binomial}(\text{Households}_l,\ 0.6)$, which is not a
convenience but the actual generative story: each household independently
either adopts the store into its routine or does not, at a common
probability, and counting independent yes or no trials is exactly the
setting in which the Binomial distribution is the maximum entropy choice.

Each customer who does adopt the store is given a weekly grocery budget,

\[
B_i \sim \text{LogNormal}(\log 85,\ 0.35),
\]

so that the median customer spends eighty five euros a week with a
realistic right tail of larger households. The LogNormal is the correct
shape for the same reason incomes
are approximately log normal in the population at large: a household's
budget is the product of several factors, namely household size, income
level, and consumption habits, and multiplying many small positive factors
produces a LogNormal distribution by the same central limit argument that
produces a Normal distribution when factors are added rather than
multiplied. The parameterization is also directly interpretable, since
$e^{\mu}$ is exactly the median and the scale $\sigma = 0.35$ places roughly
ninety five percent of customers between forty two and one hundred seventy
euros a week.

Shopping rhythm is modeled with a primary day $d_i$ drawn from a categorical
distribution weighted toward the weekend, together with two further
probabilities,

\[
\pi_i \sim \text{Beta}(9, 1.5), \qquad p_i \sim \text{Beta}(1.5, 12),
\]

where $\pi_i$ is the probability of actually shopping on the primary day
(schedule adherence) and $p_i$ is the probability of an extra, midweek top
up visit on any other day. The first piles its mass near one, so that
habits mostly hold but a minority of customers remain erratic; the second
piles its mass near zero, so that most days bring no extra visit but the
possibility is never exactly zero. In both cases the Beta distribution is
required because the object being modeled is a probability, and in both
cases the chosen parameters keep the density away from the endpoints, so
that no customer in the model is a perfectly deterministic machine.

Tastes are represented in three parts. A weekly basket composition across
the twelve merchandise categories, $w_i \sim \text{Dirichlet}(\alpha)$, is
the natural generalization of the Beta to a set of shares that must sum to
one, and its concentration vector $\alpha$ simultaneously fixes the
population's average basket, through the relative size of its entries, and
controls how much individual customers deviate from that average, through
its overall magnitude. Price sensitivity is drawn from
$\beta_i \sim \text{LogNormal}(0, 0.4)$, which must be strictly positive
because every demand curve in the model slopes downward, and whose median
of one anchors the scale of the entire utility function introduced in the
next subsection. Brand affinity, meaning the pull toward premium or budget
merchandise, is deliberately constructed to correlate with budget:

\[
b_i = \text{logistic}(0.6\, z_i + 0.8\, \nu_i), \qquad \nu_i \sim
\text{Normal}(0, 1),
\]

where $z_i$ is the customer's standardized log budget and $\nu_i$ is fresh,
independent noise. This is a standard latent Gaussian device in which an
unbounded latent trait is built with a controlled correlation and then
squashed into the unit interval with a logistic transform, and the two
loadings are chosen so that $0.6^2 + 0.8^2 = 1$, which keeps the latent
score on a standardized scale and makes the loading on budget, $0.6$,
exactly equal to the correlation between brand affinity and budget by
construction. The correlation itself is required because in reality
wealthier households do lean toward premium brands, and a dataset that
failed to reproduce it would fail a basic realism check before any analysis
began.

Finally, payment behavior is built from a persistent type rather than a
single flat probability, because a flat rate would make payment method pure
noise, uninformative about who is standing at the till:

\[
T_i \sim \text{Bernoulli}(0.6), \qquad
\text{PaysByCard}_{iv} \sim \text{Bernoulli}\big(0.95\,T_i +
0.10\,(1-T_i)\big) ,
\]

so that a card preferring customer, $T_i = 1$, pays by card on ninety five
percent of visits and a cash preferring customer on only ten percent. Card
payments carry the customer's identity forward into the receipts; cash
payments do not, which is the single mechanism that gives the eventual
dataset its realistic, partial customer level visibility.

\subsection{The discrete choice model}

The single most load bearing structure in the entire model is the rule by
which a customer, standing before a shelf of competing products within one
merchandise category, decides what to buy. Each product $s$ offers the
customer a utility

\[
U_{is} = \theta_{ic} + a_s + \gamma\left(1 - \left|b_i -
\text{BrandLevel}_s\right|\right) - \beta_i\,\text{Price}_s ,
\]

where $\theta_{ic}$ is the customer's current need for the category, $\gamma$
weights the fit between the customer's brand affinity and the product's
brand level, and $\beta_i$ is the customer's own price sensitivity. The term
$a_s \sim \text{Normal}(0, 0.8)$ is an unmodeled per-product appeal
intercept, standing in for the thousand unmodeled reasons one product
outsells its shelf neighbor, such as packaging or familiarity; a Normal is
the maximum entropy choice given only a scale, and this particular scale is
wide enough that a few products in every category sell so little all year
that they amount to realistic dead stock, a fixture of every real
assortment. Alongside every stocked product sits an
outside option, representing the choice to buy nothing here or to buy
elsewhere, and the customer's actual choice is drawn as
$\text{choice} \sim \text{Categorical}(\text{softmax}(U_{is_1}, \ldots,
U_{is_K}, U_0))$.

The softmax form is not a convenient approximation; it is the exact
consequence of a well known result in discrete choice theory. If each
product's utility is disturbed by an independent, identically distributed
Gumbel shock and the customer simply picks the option with the highest
noisy utility, the resulting choice probabilities are exactly the softmax
of the mean utilities. This is McFadden's random utility result
(\hyperlink{ref:mcfadden1974}{McFadden, 1974}), and it
means the smooth formula above and the human story of a shopper with
genuine but noisy tastes are the same model described two ways. It also
means the model the customer is implicitly following is exactly the model a
real discrete choice analyst would fit to purchase histories, so that
grading a later analytical exercise against the true parameters is a fair
comparison of like against like, rather than a comparison against a
mechanism the analyst was never given a chance to recover.

This single structure produces, without any additional machinery, four
properties that a realistic market must have. Demand slopes downward,
because raising a product's price lowers its utility and therefore its
choice probability, at a rate set by each customer's own price sensitivity.
Substitution follows a sensible pattern, because a price increase on one
product pushes probability mass onto its nearest neighbors in brand level,
not onto arbitrary alternatives. Budget discipline emerges naturally, because
a customer works through a shopping list item by item until the budget
binds. Hidden demand is recorded automatically, because when a customer's
first choice is out of stock, the original choice is logged as a latent
truth before the customer re-evaluates the softmax over what remains, giving
an honest ledger of demand that never appears in the sales data.

\subsection{Owner beliefs and the cost of opening}

The entrepreneur begins with a fixed sum of capital and three standing
policies inherited from ordinary small business practice rather than
optimized from scratch: fixed opening hours; a fixed markup $m_c$ applied
uniformly within each merchandise category, so that a product's shelf price
is set mechanically as $\text{Price}_s = \text{charm}\big((1+m_c)\,
\text{Cost}_s\big)$, with the charm function rounding to the SKU's own
habitual psychological ending, such as a price ending in nine cents; and a
flat safety stock rule requiring a constant fraction $\eta = 0.30$ of a
month's believed category demand to sit on the shelf at all times. A fourth
belief, that no single product should capture more than a share
$\rho = 0.05$ of its category's demand, rounds out the policies that the
opening decision below will actually optimize under.

Before any belief can be formed there is a fact to be believed about. The
true demand potential of a location for a merchandise category, the
quantity the whole opening decision ultimately turns on, is defined simply
as

\[
D_{cl} = \text{Households}_l \cdot \mu_c ,
\]

where $\mu_c$ is the mean monthly per household consumption of category $c$
that a store of this kind could realistically capture, already net of the
households who never adopt the store at all and of the share of spending a
neighborhood shop can expect to win against larger competitors. This
quantity is ground truth: nobody in the model, least of all the owner,
observes it directly.

The owner cannot observe the true demand potential of a location, so his
belief is modeled as the truth seen through a noisy and mildly optimistic
lens, $\hat{D}_{cl} = D_{cl}\,e^{\delta_c}$ with $\delta_c \sim
\text{Normal}(0.10, 0.25)$. The error is placed on the logarithmic scale
because people misjudge quantities in ratios rather than in absolute units;
nobody thinks their estimate is off by thirty seven boxes, they think demand
is probably a fifth higher or lower than it looks. Exponentiating a Normal
error produces a belief that is LogNormal around the truth, automatically
positive as a demand estimate must be, and the positive mean of $\delta_c$
encodes a documented behavioral regularity, namely that new business owners
tend to be mildly overoptimistic. This single design choice is what causes
the store to systematically overstock in its early period, and the size of
that overstocking can be computed exactly from the belief error rather than
being asserted as an unexplained fact about the data.

The cost of opening the store is written as an explicit structural causal
model, meaning each cost component is caused by, rather than merely equal to,
the decisions and location features that produce it:

\[
\begin{aligned}
\text{TotalCost}_0 &\overset{\leftarrow}{=} \text{FixedCost}_0 + \text{VariableCost}_0 \\
\text{FixedCost}_0 &\overset{\leftarrow}{=} \text{Rent}_0 + \text{SetupCost}_0 \\
\text{VariableCost}_0 &\overset{\leftarrow}{=} \text{RestockCost}_0 + \text{StorageCost}_0 \\
&\quad {} + \text{OverheadCost}_0 + \text{UtilityCost}_0 + \text{ListingCost}_0
\end{aligned}
\]

One detail deserves particular emphasis because it is a solvency condition
rather than a stylistic one: the owner himself draws no wage in this
accounting, since a sole proprietor is paid out of profit, not out of
payroll. A numerical audit of the model confirmed that charging the owner a
market wage for his own long working days makes every affordable opening
plan show a believed loss, which no rational entrepreneur would actually
sign, so leaving the owner unpaid is required for the model to describe a
decision anyone would make.

\subsection{The opening decision as a mixed integer program}

With beliefs, costs, and policies fixed, the owner's opening decision is
formulated as a mixed integer linear program, choosing simultaneously which
location to occupy, which products to list, how much of each to buy, and how
many additional staff to hire, so as to maximize the profit he believes he
will earn, subject to the constraints he actually knows about. Five decision
variables carry the whole problem: $y_l \in \{0,1\}$ marks which single
location is chosen, $x_s \in \{0,1\}$ marks which products are listed on the
shelf at all, $q_s \geq 0$ is the quantity of product $s$ purchased at
opening, $u_s \geq 0$ is the quantity the owner believes he will sell of it
before the objective ever needs the true answer, and $E^h_0 \geq 0$ is the
number of additional staff hired beyond the owner himself.

\[
\max \;\; \sum_{c}\sum_{s \in S_c} (1+m_c)\,\text{Cost}_s \cdot u_s \;-\; \text{TotalCost}_0
\]

subject to

\[
\begin{aligned}
\sum_l y_l &= 1 && \text{(exactly one location)} \\
u_s &\leq q_s && \text{(cannot sell what is not stocked)} \\
\sum_{s \in S_c} u_s &\leq \sum_l y_l\,\hat{D}_{cl} && \text{(believed category demand)} \\
u_s,\ q_s &\leq \rho \sum_l y_l\,\hat{D}_{cl} && \text{(no product dominates its category)} \\
\sum_{s \in S_c} q_s &\geq \eta \sum_l y_l\,\hat{D}_{cl} && \text{(the safety stock rule)} \\
\sum_s q_s &\leq \sum_l y_l\,\text{ShelfCapacity}_l && \text{(shelf capacity)} \\
\sum_s x_s &\leq \sum_l y_l\,\text{ShelfSlots}_l && \text{(shelf slots)} \\
E^h_0 &\geq \sum_l y_l\,\text{OperationalNeeds}_l - 1 && \text{(minimum staffing)} \\
\text{TotalCost}_0 &\leq \text{InitialCapital}_0 && \text{(capital)}
\end{aligned}
\]

Here $\rho = 0.05$ is the owner's flat belief that no
single product should capture more than a twentieth of its category's
demand, and $\eta = 0.30$ is the fraction of a month's believed category
demand he insists on holding as safety stock at all times, the same two
policy constants introduced at the start of this subsection.

A mixed integer linear program is the natural formulation here, rather than a
simpler heuristic, because the decision genuinely is a joint discrete and
continuous optimization: the location choice and the listing decision for
each product are binary, the purchase quantities are continuous, and every
constraint linking them, such as capital and shelf space, is linear once the
location is fixed. Solving the joint problem rather than comparing separate
solutions for each candidate location keeps the model honest about the fact
that a real entrepreneur weighs locations against each other using the very
same capital and the very same assortment logic, not eight independent plans
evaluated in isolation.

This formulation produces several properties without any of them being
separately engineered. Location choice becomes a genuine trade off between
rent and believed demand, exactly as a real entrepreneur weighs traffic
against cost. Product variety emerges from the interaction of beliefs,
capital, and the no domination constraint, rather than being a hand set
number: the breadth of the assortment is an equilibrium outcome of the
optimization, not a parameter. And because the belief error introduced above
carries a positive mean, the optimal solution systematically overstocks
relative to the truth, so that when the analyst later finds bloated inventory
in the cost sheets, there is a genuine mechanism behind it rather than an
injected anomaly.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance = 7mm and 10mm,
    box/.style = {rectangle, draw, rounded corners, align=center, minimum height=0.9cm, font=\small, text width=2.5cm},
    exo/.style = {box, fill=black!4},
    arrow/.style = {-{Latex[length=2mm]}, thick}
]
    \node[exo] (Q) {Location quality $Q_l$};
    \node[exo, below=of Q] (K) {Initial capital};
    \node[exo, below=of K] (PREF) {Customer preferences};
    \node[exo, below=of PREF] (DELTA) {Belief error $\delta_c$};

    \node[box, right=of Q] (R) {Rent, setup cost, shelf capacity};
    \node[box, below right=4mm and 18mm of PREF] (D) {True demand $D_{cl}$};
    \node[box, below=of D] (BEL) {Owner beliefs $\hat{D}_{cl}$};

    \node[box, right=26mm of R, minimum height=1.6cm] (MILP) {Opening decision\\(mixed integer program)};

    \node[box, right=of MILP] (OUT) {Location, assortment,\\quantities, staffing};

    \draw[arrow] (Q) -- (R);
    \draw[arrow] (PREF) -- (D);
    \draw[arrow] (D) -- (BEL);
    \draw[arrow] (DELTA) -- (BEL);
    \draw[arrow] (R) -- (MILP);
    \draw[arrow] (BEL) -- (MILP);
    \draw[arrow] (K) -| (MILP);
    \draw[arrow] (MILP) -- (OUT);
\end{tikzpicture}
\caption{The causal skeleton of Phase One. Everything upstream of the decision node is generative ground truth; the decision node and its consequences are the owner's boundedly rational response to it.}
\label{fig:phase1graph}
\end{figure}

\subsection{Why the owner is wrong, and why that is the point}

Figure~\ref{fig:phase1graph} summarizes the causal structure of Phase One.
Every arrow upstream of the decision node describes objective features of the
world; the decision node and everything downstream of it describes the
owner's response to a distorted and simplified picture of that world. Five
specific divergences between belief and truth are built into this phase, and
each is a documented, discoverable fact rather than an arbitrary flaw:
optimistic demand beliefs, a flat safety stock rule applied identically
regardless of season, category level markups that ignore individual price
sensitivity, static beliefs that never anticipate the seasonal cycle
introduced in the next phase, and a flat product share cap that ignores
genuinely heterogeneous tastes. Three profit figures eventually exist side by
side once the later phases are complete: what the owner believed he would
earn, what he actually earns, and what a perfectly informed operator could
have earned given the same world. The distances between these three numbers
are the quantified value of better information and better decisions, and
producing that comparison honestly, rather than asserting it, is arguably the
central purpose of the entire construction.

\section{Phase Two: The Exogenous Script of the Year}
\label{sec:phase2}

Phase One built a world frozen at a single instant. Phase Two supplies its
motion across the year that follows: the weather, the calendar, macroeconomic
shocks, and idiosyncratic household budget shocks, all decided in advance and
independent of anything the simulated agents subsequently do. This
independence is not a simplification made for convenience; it is what
licenses treating weather as a natural experiment and cost shocks as
instruments in later analysis, because nothing about the shop's own behavior
could possibly have caused the weather.

\subsection{Calendar and weather}

The simulated year runs three hundred sixty five days, aligned to a real
calendar year so that seasons, holidays, and an analyst's ordinary intuitions
all line up, including two days on which the store is legally closed.

Temperature is modeled as a deterministic annual cycle plus a persistent
stochastic anomaly, $T_t = \mu_T(t) + \varepsilon_t$, with $\mu_T(t) = 12 + 9
\cos\left(\tfrac{2\pi(t-200)}{365}\right)$ and the anomaly following a
first order autoregressive process, $\varepsilon_t = \phi\,\varepsilon_{t-1}
+ \text{Normal}(0,\sigma_\varepsilon)$ with $\phi = 0.7$. The autoregressive
form is chosen because it is the maximum entropy stationary process
consistent with exactly two facts about weather that are worth encoding and
no more: it has a stable typical spread around the seasonal mean, and warm or
cold spells persist for several days rather than resetting independently each
morning. The decomposition into a deterministic cycle and a stochastic
anomaly is not a cosmetic convenience either; it is the causal design that
allows later analysis to distinguish the predictable effect of season from
the genuinely surprising effect of weather, since the anomaly is constructed
to be orthogonal to the calendar by design, while in raw data the two are
thoroughly entangled because July is warm precisely because it is July.

Rainfall requires two separate decisions, whether it rains at all and how
much, and each has its own natural distribution. Occurrence follows a two
state Markov chain rather than an independent coin flip at each day,

\[
W_t \mid W_{t-1} \sim \text{Bernoulli}\big(p_{01}(1-W_{t-1}) +
p_{11}\,W_{t-1}\big), \qquad p_{01} = 0.25,\ \ p_{11} = 0.55 ,
\]

because weather fronts pass through and wet days genuinely cluster
together, which an independent process cannot represent; these transition
probabilities imply a stationary wet day frequency of about thirty six
percent and wet spells averaging just over two days, both ordinary for a
temperate climate. Given that it rains, the amount is drawn from a Gamma
distribution, $R_t \mid W_t{=}1 \sim \text{Gamma}(1.4, \text{mean}=5\text{
mm})$, the standard meteorological choice and the maximum entropy
distribution on the positive real line once a mean and a mean logarithm are
fixed, producing the realistic pattern of many light drizzles and few
genuine downpours.

\subsection{Category demand modifiers}

The central object of this phase is a daily, per category demand modifier
$M_{ct}$ that scales how strongly the neighborhood wants a given category of
merchandise on a given day. It is built log additively from three separate
causal channels:

\[
\log M_{ct} = a_c \cos\left(\frac{2\pi(t-200)}{365}\right) + \kappa_c\, z_t +
h_c\, H_t ,
\]

where $z_t$ is the standardized temperature anomaly and $H_t$ flags the days
immediately preceding a major holiday. The three terms are genuinely distinct
causal stories rather than three pieces of one effect. The seasonal amplitude
$a_c$ captures the predictable annual cycle, everything that makes July
differ from January whether or not the weather itself cooperates, such as
daylight and school terms. The anomaly loading $\kappa_c$ captures the pure
weather effect, an unseasonably warm week moving demand right now regardless
of what the calendar says. The holiday loading $h_c$ captures the stock up
surge in the days before a major holiday. Because the model is constructed
multiplicatively and expressed additively on the logarithmic scale, every
loading has a clean, elasticity like interpretation, and because the
temperature anomaly is orthogonal to the seasonal cycle by construction,
season and weather are separately identifiable in the ground truth, which
lets a later diagnostic exercise distinguish a genuine weather effect from a
confounded seasonal one.

One category is deliberately given all zero loadings, meaning nothing in
preferences makes it more or less wanted at any time of year. Its observed
sales are nevertheless not perfectly flat in the resulting data, because
every category draws from the same household budget, and when other
categories surge seasonally they crowd the flat category out of a tight
weekly budget. This produces a subtler lesson than a simple negative control:
an analyst who finds an apparent seasonal pattern in this category and stops
there has mistaken a budget mediation effect for a genuine preference effect,
and only tracing the mechanism reveals the difference.

A single loading per category would, however, average away the most
recognizable seasonal pattern in retail, namely that demand for products such
as ice cream is dramatically higher in summer, a pattern that would otherwise
be diluted by averaging it together with other, flatter products in the same
broad category. Rather than assigning a separate seasonal loading to every
narrow product type, which would require an unmanageable number of
essentially arbitrary parameters, the model adds a narrow, surgical
refinement: a small number of specific product types whose seasonality
flagrantly contradicts their category's average receive their own tilt term,

\[
\log \psi_{pt} = a_p \cos\left(\frac{2\pi(t-200)}{365}\right) + \kappa_p\,
z_t ,
\]

built exactly like the category modifier above but at the finer grain of a
product type $p$ within a category, and entering the same choice model
introduced in Phase One as an additional utility component,
$U_{ist} = U_{is} + \log\tilde{\psi}_{p(s),t}$. In winter, a customer's need
for the broader category therefore slides from the summer leaning product
toward its winter leaning neighbor within the same shopping decision,
without any new machinery beyond the existing discrete choice structure;
every other product type keeps $\psi \equiv 1$.

\subsection{Foot traffic and household budgets}

Weather does not only change what people buy; it changes whether they come
to the store at all. A daily traffic modifier

\[
\log \Lambda_t = -0.20\,W_t + 0.15\,H_t + \varepsilon^{\Lambda}_t, \qquad
\varepsilon^{\Lambda}_t = \phi^{\Lambda}\,\varepsilon^{\Lambda}_{t-1} +
\text{Normal}(0, \sigma^{\Lambda}_\varepsilon), \qquad \phi^\Lambda = 0.85 ,
\]

scales every customer's visit probability downward by about eighteen
percent on a wet day and upward by about sixteen percent in the days before
a major holiday, with the same autoregressive logic as the temperature
anomaly so that a slow week tends to stay slow for a few days rather than
resetting each morning. A subtle but important consequence follows automatically from
the pantry mechanism introduced in Phase Three: because a customer's need
for a category grows with time since their last purchase, a customer rained
off on one day simply arrives hungrier a few days later. The transaction
data therefore shows the familiar retail pattern of a dip on the rainy day
followed by a rebound shortly after, purely as an emergent consequence of two
mechanisms that never reference each other, with no separate rebound
parameter anywhere in the model.

Household budgets are not held perfectly steady across the year either.
Each week's budget is built as

\[
B_{iw} = B_i \cdot e^{\xi_{iw}} \cdot s_{iw} \cdot g_{iw}, \qquad
\xi_{iw} \sim \text{Normal}(0, 0.12) ,
\]

where the wobble $\xi_{iw}$ is a small multiplicative perturbation
representing ordinary week to week circumstance, on the same logarithmic
scale as the budget itself, and $s_{iw}$ and $g_{iw}$ encode two rarer,
more consequential mechanisms. A tight spell sets $s_{iw} = 0.65$ for its
duration, representing a genuine financial hardship such as a job loss or
an unexpected repair; it begins with probability
$p_w = 0.004\,(1 + 2\,\bar{g}_w)$ per week, where $\bar{g}_w \in [0,1]$ is
the scripted energy crisis's own trajectory described below, and persists
for $\text{Geometric}(1/6)$ further weeks, a memoryless duration averaging
six weeks. A splurge sets $g_{iw} = 1.5$ for a single week with probability
$0.01$, representing a windfall or a celebration. The tight spell's entry
rate is deliberately coupled to the crisis trajectory, tripling at the
crisis's peak, because a real energy crisis squeezes household heating
bills at the same time as it squeezes a retailer's refrigeration costs, and
a model that raised the store's costs without touching any household's
budget would be describing an implausible economy.

\subsection{Cost shocks as a marked point process}

Beneath the everyday weather sits a slower moving layer of macroeconomic
shocks affecting wholesale costs, and underneath that a boring, continuous
inflation drift, $\text{infl}(t) = e^{\pi t/365}$ with $\pi = 0.025$ a year,
riding every cost and rate in the model. The shocks proper are modeled as a
marked point process: shock events arrive according to a Poisson process,

\[
K \sim \text{Poisson}(1.5), \qquad \tau_k \sim \text{Uniform}(1, 365)\ \text{for each event } k ,
\]

since the Poisson count is the maximum entropy distribution for the number
of rare, independent events given only a rate, and a well known property of
the Poisson process guarantees that, given the count, the arrival times are
conditionally uniform across the year, which is a theorem rather than an
additional assumption. One large event, an autumn energy crisis, is
scripted rather than drawn at random, ensuring that every run of the model
tells a genuinely interesting supply side story rather than leaving open
the possibility of a year with no shocks at all.

Each event's magnitude is drawn from a LogNormal distribution,

\[
\zeta_k \sim \text{LogNormal}(\log \zeta_0,\ 0.3),
\]

because a shock's peak effect is a positive multiplicative disturbance on
cost, the same support and story argument that justified every other ratio
like quantity in the model, and each event's effect over time follows a
deterministic shape,

\[
g_k(t) = \zeta_k \cdot \min\!\left(1, \frac{t-\tau_k}{14}\right) \cdot
\exp\!\left(-\frac{(t-\tau_k-14)_+}{\rho_k}\right), \qquad t \geq \tau_k ,
\]

ramping up linearly over two weeks and then decaying exponentially with
time constant $\rho_k$, so that a shock is neither a permanent level change
nor a transient blip but a slow wave that a forecaster genuinely has to
track. Every affected category's wholesale cost multiplier then assembles
the drift and every live shock log additively,

\[
\log \text{CostMult}_{ct} = \frac{\pi t}{365} + \sum_k g_k(t)\,
\mathbb{1}[c \in A_k] ,
\]

with $A_k$ the set of categories the event touches.

The mechanism by which these shocks reach the shelf is one of the more
consequential design decisions in the whole model. The owner never observes
$\text{CostMult}_{ct}$ directly; he sees only the invoices his supplier
sends him, and he tracks an exponentially smoothed trend in those invoice
costs,

\[
\overline{\text{Cost}}_{s,t} = \alpha\,\text{Cost}_{s,t} + (1-\alpha)\,
\overline{\text{Cost}}_{s,t^-}, \qquad \alpha = 0.35 ,
\]

where the smoothing weight $\alpha = 0.35$ is a purely technical constant
unrelated to the basket concentration vector of the same name in
Section~\ref{sec:phase1}. The owner moves the shelf tag only once the
smoothed cost has drifted more than three percent from the current price, a
pattern known in the economics
literature as menu cost pricing. The consequence is that shelf prices now
vary over time for reasons entirely unconnected to the demand side of the
market: a foreign harvest failure or a port strike moves costs without
touching any household's budget, which is exactly the exogenous price
variation an econometrician needs in order to estimate a genuine price
elasticity of demand rather than a number contaminated by reverse causation
from demand to price. The scripted energy crisis is a deliberate exception
to this cleanliness, since it also squeezes household budgets through the
tight spell mechanism described above, making it a subtly invalid instrument
whose bias can, in principle, be computed exactly, turning instrument
validity from an assumption asserted in a single sentence into a testable
claim with a known answer.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance = 6mm and 8mm,
    box/.style = {rectangle, draw, rounded corners, align=center, minimum height=0.85cm, font=\small, text width=2.5cm},
    exo/.style = {box, fill=black!4},
    arrow/.style = {-{Latex[length=2mm]}, thick}
]
    \node[exo] (SEA) {Annual cycle};
    \node[exo, below=of SEA] (ANOM) {Temperature anomaly};
    \node[exo, below=of ANOM] (RAIN) {Rain occurrence and amount};
    \node[exo, below=of RAIN] (HOL) {Holiday calendar};
    \node[exo, below=of HOL] (EVT) {Macro shock events};

    \node[box, right=22mm of ANOM] (M) {Demand modifiers $M_{ct}$};
    \node[box, below=of M] (LAM) {Traffic modifier $\Lambda_t$};
    \node[box, below=of LAM] (CM) {Wholesale cost paths};
    \node[box, below=of CM] (BW) {Household budgets};

    \node[box, right=26mm of LAM, minimum height=1.6cm] (P3) {Phase 3\\daily market};

    \draw[arrow] (SEA) -- (M);
    \draw[arrow] (ANOM) -- (M);
    \draw[arrow] (HOL) -- (M);
    \draw[arrow] (RAIN) -- (LAM);
    \draw[arrow] (HOL) -- (LAM);
    \draw[arrow] (EVT) -- (CM);
    \draw[arrow] (EVT) -- (BW);

    \draw[arrow] (M) -- (P3);
    \draw[arrow] (LAM) -- (P3);
    \draw[arrow] (CM) -- (P3);
    \draw[arrow] (BW) -- (P3);
\end{tikzpicture}
\caption{The causal skeleton of Phase Two. Every arrow points into the market that Phase Three performs; nothing points back, which is what licenses treating these channels as exogenous instruments and natural experiments.}
\label{fig:phase2graph}
\end{figure}

\subsection{Why exogeneity licenses counterfactual reasoning}

Because Phase Two's script is written entirely in advance and never listens
to anything the simulated agents do, a counterfactual world can be produced
almost for free: simply edit the script, for instance deleting the scripted
energy crisis, and replay the identical daily market mechanism of Phase Three
with every other random draw held fixed by the keyed randomness discipline
described in Section~\ref{sec:principles}. The difference between the
original run and the edited run is then the causal effect of the edit
itself, not merely an estimate of it. This is the property that Phase Four,
described in Section~\ref{sec:phase4}, generalizes into a full policy
laboratory.

\section{Phase Three: The Daily Market Loop}
\label{sec:phase3}

Phase Three is where the two sides of the market, built and scripted in the
preceding phases, finally meet. It is the only phase that produces data in
the ordinary sense: a customer with a specific need walks past a shelf whose
price was set according to the owner's rule, on a day whose weather was
drawn in Phase Two, and either buys, substitutes, or leaves, and every one of
these outcomes either does or pointedly does not leave a trace in the
records the analyst eventually receives.

\subsection{The pantry and the mechanics of need}

Phase One promised that a customer's need for a category would be driven by
their basket preferences and the time since their last purchase. Phase Three
makes that promise mechanical through the simplest object that behaves
correctly: a pantry. Each customer is given a latent stock of each category,
measured in units, which drains at a steady daily rate and refills only when
the customer actually buys more:

\[
P_{ic,t+1} \overset{\leftarrow}{=} \max(0,\ P_{ict} - r_{ict}) +
(\text{units purchased on day } t), \qquad r_{ict} = r_{ic}\cdot M_{ct} .
\]

Scaling the drain rate itself by the demand modifier, rather than adding the
modifier directly to a customer's momentary desire, is a deliberate
refinement of the interface promised in Phase Two, because it changes not
only how much a customer wants a category but when that want actually
arrives: a hot week does not make a household want ice cream in the
abstract, it makes the household eat the ice cream in the freezer faster,
so the container runs out sooner and the shopping trip that replaces it
arrives earlier. This single change is also what makes the rain rebound
described in the previous section emerge with a realistic magnitude, since a
missed shopping trip simply leaves the drain running against an unreplenished
pantry.

A customer's need for a category is then defined as the shortfall of their
current pantry stock against a target level representing several days of
comfortable cover,

\[
\theta_{ict} = \alpha\,\frac{\max(0,\ P^{\text{tgt}}_{ic} - P_{ict})}
{P^{\text{tgt}}_{ic}}, \qquad P^{\text{tgt}}_{ic} = \tau\, r_{ic}, \qquad
\tau = 10 \text{ days},
\]

with the shape parameter $\alpha$, again a purely local constant unrelated
to the two earlier uses of the same letter, calibrated so that a fully empty pantry
presses harder on the choice model than an ordinary price difference but
not harder than an extreme one. This shortfall enters the utility function
of Phase One's discrete choice model directly, exactly as $\theta_{ic}$ did
there. An empty pantry presses hard on the
customer's decisions; a full one leaves the category to the outside option
entirely, meaning the customer simply does not shop for it this trip. One
bookkeeping rule is essential to the model's long run stability: when the
outside option wins, meaning the customer effectively buys the item
elsewhere or does without, the pantry is nonetheless refilled, with the
event recorded in a hidden ledger under its own cause. Without this rule,
demand lost to competitors would accumulate without bound and every basket
would eventually list every category; with it, the pantry remains stationary
and the store's captured share of total demand is exactly what the discrete
choice model implies it should be.

\subsection{Building the basket and the hidden demand ledger}

A visiting customer builds a shopping list of every category whose pantry
cover has fallen below a replenishment threshold, walks the list in
descending order of urgency, and at each category faces the same discrete
choice model introduced in Phase One, now evaluated only over products
currently in stock, with any live promotional discount folding directly into
the effective price the customer sees. Because a discount is treated simply
as a lower price, promotional response requires no separate machinery and
automatically inherits each customer's own individual price sensitivity.
Having chosen a product, the customer buys toward the shortfall rather than
filling it exactly,

\[
\text{Quantity} = \Big\lceil \big(P^{\text{tgt}}_{ic} - P_{ict}\big)\,
\kappa \Big\rceil, \qquad \kappa \sim \text{Uniform}(0.7, 1.0) ,
\]

capped by a per trip carrying limit drawn uniformly between eight and
fifteen units and by whatever remains of the week's budget, so that
quantity is a direct consequence of the pantry rather than a freshly drawn
random object, which is what gives the model its realistic correlation
between a longer gap since the last visit and a larger basket this time.

When a customer's chosen product is out of stock, three things happen in a
fixed order. First, the original intended choice is logged as a latent
truth, with its cause recorded as a stockout. Second, the customer
re-evaluates the same discrete choice model with the unavailable product
removed from the choice set, so that probability mass flows toward the
nearest neighbors in brand level, exactly the substitution pattern the
original choice model was built to produce. Third, if the re-evaluated
choice lands on another available product, the sale proceeds; if it lands on
the outside option instead, the store has lost a sale it will never observe
in its own data. Four distinct causes of invisible demand accumulate in this
way over the year: an arrival outside opening hours, a stockout, an
exhausted weekly budget partway through the shopping list, and the outside
option simply winning on price or availability. This hidden demand ledger is
the honest, mechanistic version of the claim that sales data are demand data
censored in several distinct ways, each with its own recorded cause, rather
than an unexplained gap between two numbers.

\subsection{The owner's adaptive policies}

Unlike the exogenous script of Phase Two, the owner in Phase Three reacts to
what he actually observes, and what he observes is never the full truth.
Once a week the owner places a restocking order, sized according to the
simplest forecast a real shopkeeper actually runs in practice: a trailing
four week average of his own recorded sales,

\[
\hat{D}^{\text{op}}_{cw} \overset{\leftarrow}{=} \frac{1}{4}
\sum_{j=1}^{4} \text{Sales}_{c,w-j} .
\]

He then orders up to a target sized for the review and lead time ahead of
him, plus his standing safety stock rule,

\[
\text{Target}_{sw} = \left(\frac{R+L}{7} + \eta\,\frac{30}{7}\right)
\hat{D}^{\text{op}}_{cw}\,\tilde{\sigma}_{sw} ,
\]

with $R = 7$ days between reviews, $L = 2$ days of lead time, and
$\tilde{\sigma}_{sw}$ the product's recent share of its category's sales,
smoothed with one pseudo unit so that no listed product is ever starved to
zero. This rule carries two distinct, realistic flaws built directly into
its structure. It is backward looking, so that a rising seasonal trend is
always underestimated by an average of the weeks before it rose, and the
month stale allocation $\tilde{\sigma}_{sw}$ starves whichever product's
season is arriving while fattening the one whose season just left, even
while the category total stays roughly on track. And it is blind to
censoring, because the sales it averages are themselves truncated by the
store's own past stockouts, which produces a self-reinforcing pattern
sometimes called a censoring spiral: a product that stocks out repeatedly
reports weak sales, earns a smaller share of the next order, and
consequently stocks out again. Both flaws are genuine, previously
documented pathologies of naive forecasting, reproduced here as the
mechanical consequence of an otherwise ordinary and defensible business
rule, rather than injected as an arbitrary model defect.

The owner also runs two distinct kinds of promotion, and the contrast
between them is itself an intended lesson. A markdown campaign is triggered
endogenously whenever a category's weeks of cover on hand exceeds a
threshold, targeting specifically the slowest selling third of that
category's products. Because the trigger is overstock, and overstock is
correlated with weak underlying demand, any naive before and after
comparison of a markdown's effect is contaminated by exactly the same
selection problem that afflicts real world marketing analytics, understating
the campaign's true effect because the campaign was deliberately aimed at
products already trending downward. A separate, calendar triggered
storewide discount on the last Sunday of every month is, by contrast,
entirely exogenous, announced in advance and correlated with nothing about
current stock levels, providing a clean source of price variation an
analyst can use to anchor an elasticity estimate before attempting to
untangle the confounded markdown campaigns.

Perishable products spoil overnight,

\[
\text{Spoiled}_{st} \sim \text{Binomial}\big(\text{OnHand}_{st},\
\min(0.35,\ \lambda^d_{c(s)}\,f_{c(s),t})\big), \qquad
\lambda^d_c = 1 - (1-\lambda_c)^{1/7} ,
\]

where $\lambda_c$ is the category's interpretable weekly base rate,
converted to a nightly rate $\lambda^d_c$, and modulated by a spoilage
factor built from the same log additive logic as every other multiplier in
the model,

\[
\log f_{ct} = \kappa^T_c\,\frac{T_t - \bar{T}}{10} + \kappa^E_c\,\bar{g}_t
+ \text{Normal}(0, 0.25) ,
\]

where $\bar{T}$ is the annual mean temperature, so the heat term is a
deviation from typical conditions rather than an absolute temperature, with
a heat term strongest for ambient perishables such as bread and
produce, a cold chain strain term loading on the energy crisis's own
trajectory $\bar{g}_t$ and strongest for frozen and dairy goods, and an
idiosyncratic batch luck term standing in for the reality that one delivery
travels worse than another. The thinning itself uses a nightly binomial
process rather than a single weekly cull, specifically because the store's
inventory snapshots are visible day by day, and a once weekly cull would
stamp an artificial sawtooth pattern into every perishable product's
inventory series, a periodic artifact with no physical cause that a careful
analyst would immediately, and correctly, flag as evidence of a simulated
rather than a found dataset.

Finally, the store's cash position is tracked explicitly, with a credit
line available at a fixed interest rate whenever the monthly close would
otherwise take cash below zero. Restocking orders are themselves capped by
available cash and credit headroom, so that a struggling store visibly
thins its own shelves months before any trouble would show up in an annual
profit figure, exactly as a real distressed retailer's inventory often
signals difficulty well before its accounts do.

\subsection{The recording layer: imperfect paperwork over a perfect physical reality}

An early, fully reconciled version of the generated dataset was reviewed and
found, on inspection, to have one serious flaw: its records were perfect.
Revenue tied to the cent, every inventory identity closed exactly, and no
document was ever missing, duplicated, or mistyped. Real bookkeeping is never
like this, and a dataset that reconciled perfectly would itself be evidence
of fabrication to any experienced reader. The model therefore adds a
deliberate recording layer, governed by one guiding principle: the
underlying physical reality of the simulation remains untouched and
internally perfect, and only the documents describing that reality are
allowed to err, with every imperfection injected at the point of export
rather than into the simulation's own state.

A small number of realistic paperwork defects are injected at low, carefully
calibrated rates, including duplicated till receipts from a point of sale
terminal retrying an upload, invoices posted twice by a distracted clerk,
occasional missing invoice lines for goods that were paid for but never
logged, unlogged spoilage that only surfaces later as an unexplained monthly
shrinkage, and the occasional mistyped inventory count that self corrects
the following night. Each defect is drawn from its own independently keyed
random stream and is simultaneously written to a separate answer key
recording exactly what was changed, so that the exercise of reconciling the
messy records back to the truth is itself gradable against a known,
recorded answer, and the underlying simulated reality that the counterfactual
machinery of Phase Four depends on is never disturbed by paperwork noise
that exists only in the documents an analyst receives.

The paperwork drift this produces has its own exact algebra. Let $D_{st}$
denote SKU $s$'s cumulative book-minus-true discrepancy on day $t$, pushed
upward by duplicate invoices and unrecorded spoilage and pushed downward by
missing invoices and duplicate receipts. The inventory figure an analyst
actually sees is book stock rather than the simulation's true stock $S_{st}$,

\[
B_{st} = S_{st} + D_{st} - \sum_{c \le t} A_{sc},
\]

where $A_{sc}$ is the cumulative correction posted at each monthly count. On
the last day of every month the owner counts the shelf and posts the
adjustment $A_{sc} = (B - S)_{sc^-}$ under the write-off reason
\texttt{stock\_count}, snapping the book back to physical truth before it
begins drifting anew; every other write-off keeps the reason
\texttt{spoilage}. Because unrecorded spoilage is the dominant contributor to
$D_{st}$, these monthly adjustments are themselves mostly losses, reproducing
the familiar shape of retail shrinkage, with occasional gains where a missing
invoice or a duplicated receipt happened to tip the balance the other way.
Genuine refunds are deliberately kept out of this recording layer and
modeled instead as a true simulation event: after each receipt is finalized,
an independently keyed stream determines whether one of its lines is
returned,

\[
\Pr(\text{refund} \mid \text{receipt}) = 0.006, \qquad \text{return delay} \sim
\text{Uniform}\{1, \dots, 13\}\ \text{days},
\]

posting as its own till transaction at the price actually paid, referencing
the original receipt, and destroying the returned unit rather than
restocking it, exactly as Section~\ref{sec:accounting} describes.

\subsection{The three profit figures}

With a full year performed, three profit figures that Phase One promised
finally become computable side by side: the profit the owner believed he
would earn at the moment of opening, $\pi_{\text{believed}}$, read directly
off the Phase One mixed integer program's own objective value and extended
by his own rules over the year; the profit the ledger actually shows after a
full year of realistic behavior, $\pi_{\text{realized}}$, the number a real
accountant would certify; and the profit an otherwise identical but
perfectly informed operator could have earned given the same world, meaning
the same customers, the same script, and the same policy shapes, but with
true demand in place of the owner's censored, backward looking forecast,
$\pi_{\text{oracle}}$. The two gaps between these figures are exact
differences between fully computed runs of the same mechanism rather than
estimates,

\[
\Delta_{\text{optimism}} = \pi_{\text{realized}} - \pi_{\text{believed}},
\qquad
\Delta_{\text{analytics}} = \pi_{\text{oracle}} - \pi_{\text{realized}},
\]

where $\Delta_{\text{optimism}}$ measures the cost of the owner's initial
overoptimism and $\Delta_{\text{analytics}}$ measures the value of better
information and better decisions, which is the single number the entire
construction exists to make credible by actually computing it rather than
merely asserting a plausible size for it.

\section{Phase Four: The Policy Laboratory}
\label{sec:phase4}

Phases One through Three together produce one performed year of a single
world. Phase Four turns that single performance into a laboratory in which
the same world can be re-performed under a deliberately edited set of macro
conditions, so that the causal effect of a policy change or an external
shock can be measured directly, as an exact difference between two runs of
an otherwise identical mechanism, rather than estimated statistically from a
single realized history.

\subsection{Scenarios as edited scripts}

A scenario, in this model, is nothing more than a declarative set of edits to
the exogenous script of Phase Two or to one of the owner's policy knobs, such
as his opening hours or his staffing level. Concretely, a scenario spec edits
one or more named quantities the earlier sections already defined: extra
cost events riding the same trajectory $g_k(t)$ of
Section~\ref{sec:phase2}, a multiplier on the daily traffic modifier
$\Lambda_t$, a multiplier on the weekly household budget $B_{iw}$, or a
new rate path $r_c(t)$ from a given date. Critically, a scenario never
edits an underlying random number stream directly. Because every draw in the
model is keyed by stable identity, as described in
Section~\ref{sec:principles}, replaying an edited script through the
identical daily market mechanism reproduces every unaffected customer's
decisions bit for bit, so that the entire difference between the baseline
world and the edited one is attributable to the edit by construction. This
is the same common random numbers discipline introduced to support the
Phase Two counterfactuals, here generalized into a standing feature of the
model rather than a special case. A handful of reference scenarios
illustrate the range of questions this laboratory can answer, including a
broad economy wide supply shock, a short severe weather event, a change in
value added tax, a temporary household income transfer, and a staffing
decision to hire an additional clerk and extend opening hours.

The owner's declared indifference to tax strategy is not merely asserted but
is, for the profit tax specifically, provably harmless: because
$\text{ProfitTax}$ is a flat rate on a positive base,
maximizing $(1-r^{\pi})\,\pi$ over any set of operating decisions has
exactly the same maximizer as maximizing $\pi$ itself,

\[
\arg\max_{\text{decisions}}\ (1 - r^{\pi})\,\pi(\text{decisions}) =
\arg\max_{\text{decisions}}\ \pi(\text{decisions}),
\]

so an owner who ignores the profit tax when making ordinary operating
decisions loses nothing relative to one who optimizes around it. No such
guarantee holds for the value added tax, whose category specific rate
$r_c(t)$ changes relative prices across categories and therefore can, in
principle, reward cross category arbitrage; the owner's failure to perform
that arbitrage is exactly the gap a later prescriptive analysis can
measure against a scenario's twin arms.

\subsection{The tax layer}

Because a policy laboratory without taxation would be unable to represent
the most common lever through which real macroeconomic policy actually
reaches individual households and businesses, this phase also introduces a
complete, differentiated tax system into the baseline world itself, rather
than reserving it only for edited scenarios. Value added tax is applied at
a category specific rate $r_c(t)$, and the owner's shelf tag is formed from
the tax exclusive, smoothed invoice cost with the current rate reapplied on
top:

\[
\text{Tag}_{st} = \text{charm}\Big(\overline{\text{NetCost}}_{s,t} \cdot
(1+m_c) \cdot (1+r_c(t))\Big) ,
\]

where $\overline{\text{NetCost}}_{s,t}$ is the exponentially smoothed,
tax exclusive invoice cost introduced in Phase Two and $m_c$ is the same
category markup fixed in Phase One. Because the rate enters multiplicatively
on the same trend the owner already tracks, a rate change reaches his
\emph{intended} price immediately and exactly, while the price customers
actually see at the shelf continues to lag through the same menu cost
mechanism described in Phase Two, which is precisely the kind of
incomplete, delayed tax pass-through documented in the empirical public
finance literature.

Value added tax is remitted monthly as the difference between output tax
collected on sales and input tax paid on purchases,

\[
\text{VAT}_t = \sum_c \text{Sales}_{ct}\,\frac{r_c}{1+r_c} \;-\;
\sum_c \text{Purchases}_{ct}\,\frac{r_c}{1+r_c} ,
\]

and this net amount leaves the till at the monthly close, the same till
that funds the cash capped ordering rule of Phase Three. Payroll tax applies
to any wages paid to hired staff at a flat employer rate,
$\text{PayrollTax}_t = r^{\text{pay}}\,\text{Wages}_t$ with
$r^{\text{pay}} = 0.25$, and a flat tax applies to positive annual profit,
accrued in the year it is earned rather than paid immediately,

\[
\text{ProfitTax} = r^{\pi} \cdot \max(0,\ \text{ProfitBeforeTax}), \qquad
r^{\pi} = 0.20 .
\]

The owner is modeled as tax aware but not tax strategic: he correctly
prices on his net of tax margin, as the tag formula above shows, and
correctly accounts for the full employer cost of hiring, meaning wages
plus payroll tax, in the opening decision of Phase One, but he never
performs cross category tax arbitrage or elasticity aware repricing in
response to a tax change, leaving that gap as an open door for a later
prescriptive analysis to identify and price.

\section{Phase Five: Three Years of Growth, Churn, and Capital}
\label{sec:phase5}

A single simulated year is, in one specific and important sense, an
incomplete dataset: it cannot separate a genuine growth trend from ordinary
seasonality, it cannot show an individual customer arriving in or departing
from the neighborhood, and its owner never faces the central question every
real small business eventually faces, namely whether to reinvest, hold
steady, or exit. Phase Five extends the horizon to three full years and adds
exactly the three kinds of motion a single year structurally cannot contain.

\subsection{The customer panel as a flow}

Where earlier phases treated the customer population as fixed for the
simulated year, Phase Five gives each regular customer a persistence type
drawn once at creation,

\[
\text{type}_i =
\begin{cases}
\text{rooted} & \text{with probability } 0.80 \\
\text{transient} & \text{with probability } 0.20 ,
\end{cases}
\]

where a rooted customer will not leave the neighborhood within the three
year window at all, and a transient customer instead faces an ongoing
monthly hazard of departure, $h = 1/18$, giving a mean tenure of eighteen
months. Every departure is matched, after a random delay of
$1 + \text{Geometric}(0.5)$ months, by a fresh customer profile representing
a new household moving into the vacated home, and a slow trickle of
genuinely new arrivals, $\text{Poisson}(4)$ per year, on top of
straightforward replacement represents modest organic neighborhood growth.
Newly arrived customers are themselves drawn with an even fifty percent
chance of either persistence type, reflecting the reasonable assumption
that a household which has just moved once is more likely to move again
than the settled population it replaces.

Nothing about this churn is announced anywhere in the visible data; a
customer's card token simply stops appearing in the receipts one month and
never returns, and a new token begins appearing elsewhere. An analyst must
infer departure from silence, and must take care not to confuse a customer
who has genuinely left the neighborhood with one who is merely visiting less
often for other reasons, a distinction that becomes especially difficult
near the very end of the observed window, where a recently silent customer
cannot yet be distinguished from one who will return next month. This is a
form of the right censoring problem familiar from survival analysis, and it
is a genuine, unavoidable feature of any panel observed for a finite length
of time, reproduced here rather than glossed over.

\subsection{Retained earnings and the endogenous expansion decision}

From the beginning of the second year, once the owner's books have
formalized, he begins retaining a share $\rho_{RE} = 0.5$ of every
profitable month's after tax result $\pi_m$ as declared retained earnings,
drawing the rest out as his own income,

\[
\text{draw}_m = (1-\rho_{RE})\,\max(0, \pi_m), \qquad
\text{RE}_m = \text{RE}_{m-1} + \rho_{RE}\,\max(0,\pi_m) - \text{capex}_m ,
\]

with no draw and no retention at all in a loss month, exactly as a small
business owner who has proven a concept typically begins to behave. Once
retained earnings and available cash both cross fixed thresholds,

\[
\text{RE}_m \geq K_{\text{exp}} \quad \text{and} \quad
\text{cash}_m - \text{tax\_reserve}_m \geq C_{\text{exp}} ,
\]

the owner expands the business on the first day of the following month: he
hires his first employee on a fixed part time shift, extends opening hours
by two hours a day, deepens his shelf capacity without widening the range
of products carried, and pays a one time fit out cost $C_{\text{exp}}$ from
savings.

The threshold $K_{\text{exp}}$ is calibrated, rather than tied to a fixed
calendar date, so that under the realized path of the model the expansion
lands in the autumn of the second year, after a scripted local population
increase but before a scripted competitive entry in the following year,
producing a specific and
analytically important narrative: the expansion looks like an entirely
sensible decision at the moment it is made, based on everything the owner
could reasonably have known at the time, and only becomes questionable in
hindsight once a subsequent, unrelated shock arrives. A counterfactual twin
of the model, in which the expansion threshold is disabled entirely, reveals
that the expansion's true cost across its first fourteen months substantially
exceeds the additional margin the extra hours and capacity actually generate,
because the fixed cost of an additional wage is paid in full while the
additional revenue it produces is diluted by the ordinary grocery margin.
This is a deliberately reproduced version of one of the most common capital
allocation mistakes in small retail, namely investing capital at what
retrospectively turns out to be the peak of the underlying opportunity,
generated here from the interaction of an ordinary retention rule, an
ordinary calibrated threshold, and an otherwise unrelated later shock, rather
than scripted as an explicit narrative choice.

\subsection{A world that stays unquiet}

Consistent with the stated design principle that a three year history should
not read as a single stable year merely repeated three times, the second and
third years are populated with a specific sequence of scripted events, each
chosen to test a distinct analytical skill. The second year includes an
equipment failure that destroys a portion of frozen and refrigerated stock
overnight, a single category cost shock affecting dairy products, a
prolonged summer heatwave that simultaneously boosts demand for
warm-weather products and increases spoilage pressure through the same
mechanism, and the scripted population increase that helps trigger the
expansion described above. The third year includes a contractual rent
increase taking effect at a two year lease renewal, alongside similarly
lumpy, dated steps in wages and utility tariffs, each a discrete jump on a
known date rather than a smooth index, so that an analyst who treats every
cost as continuously drifting inflation will misdate exactly where the
squeeze came from. It also includes the entry of a discounting competitor a
short distance away whose effect on individual customers is deliberately
heterogeneous,

\[
m_i = \exp(-\chi \cdot \tilde{s}_i \cdot k_i), \qquad k_i =
\begin{cases} 1.0 & \text{transient customers} \\ 0.6 & \text{rooted customers,} \end{cases}
\]

a multiplicative penalty on each customer's visit probability, where $m_i$
denotes this penalty and is unrelated to the category markup $m_c$ of
Section~\ref{sec:phase1} despite the shared letter, $\tilde{s}_i$ is that
customer's percentile rank in price sensitivity, and
$\chi$ is calibrated to a target aggregate visit drop, so that the
competitor draws away price sensitive customers fastest, and rooted
customers, held by relationships rather than price alone, are penalized at
only sixty percent of the rate a transient customer at the same price
sensitivity would be. This is followed by a scheduled rather than reactive
pricing response by the owner to that competitor, a short seasonal street
festival, and a second cost shock affecting staple goods. Two further
counterfactual twins, one in which the competitor never enters and one in
which the expansion never occurs, allow the true cost of each event to be
measured directly rather than inferred, and a specific, intended finding of
the model is that the capital decision to expand costs the business
substantially more than the competitive entry does, even though the
competitor is the more visible and more emotionally salient event to an
owner living through it.

\section{The Accounting Layer: Reconciling Money and Goods}
\label{sec:accounting}

Running beneath every phase described above is a strict accounting
discipline governing how money and physical goods move between the
artifacts an analyst eventually receives. One principle governs all of it:
the owner's own monthly ledger is always correct about money, and every
other document, particularly the raw till receipts and the raw supplier
invoices, is the analyst's paperwork to reconcile against it, complete with
the deliberate imperfections described in Section~\ref{sec:phase3}.

Revenue is recognized on a cash basis at the moment of sale, with the
effective price paid reflecting, in order, the current shelf tag, any live
promotional markdown, and the additional monthly storewide discount when it
applies. A small share of receipts contain a voided mis-ring, in which a
cashier corrects an immediate scanning error, netting to zero money and zero
goods, and are distinguishable from genuine refunds by the absence of a
reference to an earlier receipt. Genuine refunds, occurring at a low but
nonzero rate, post as an independent till transaction referencing the
original sale, reduce both cash and revenue on the day of the return, and
destroy rather than restock the returned item, since a small grocer does not
typically resell a returned perishable good.

Inventory is tracked as book stock rather than a continuous physical count,
governed by a strict identity in which each day's change in recorded stock
equals recorded deliveries minus recorded sales minus recorded write-offs,

\[
\text{book}(t) - \text{book}(t-1) = \text{deliveries}(t) - \text{sales}(t) -
\text{write\_offs}(t),
\]

with any deviation from this identity traceable either to the deliberate
snapshot typos of the recording layer or, once a month, to a physical stock
count that reconciles the books and posts the difference as a shrinkage
adjustment. Perishable spoilage and monthly stock count corrections are kept
in strictly separate categories precisely because they measure two
different things, namely genuine physical waste and pure bookkeeping drift,
and conflating them would double count the same euros.

The monthly ledger itself aggregates revenue, the cost of goods procured,
rent, wages, utilities, storage, promotional costs, net tax remittance, and
credit interest into a single bottom line, with a credit facility drawn
automatically whenever the month's closing cash position would otherwise
fall below zero. At the annual horizon this bottom line is made explicit as

\begin{align*}
\text{ProfitBeforeTax} \overset{\leftarrow}{=}\ & \sum \text{Revenue} -
\sum\big(\text{Procurement} + \text{Rent} + \text{Wages}\big) \\
&\quad {} - \sum\big(\text{PayrollTax} + \text{Utilities} +
\text{Storage}\big) \\
&\quad {} - \sum\big(\text{Flyers} + \text{VAT} +
\text{CreditInterest}\big) - \text{Openings},
\end{align*}

where $\text{Openings}$ collects any one time capital expenditure, such as a
second location's fit-out cost, incurred that year. Profit tax is then levied
only on a genuine surplus,

\begin{align*}
\text{ProfitTax} &= 0.20 \times \max(0, \text{ProfitBeforeTax}), \\
\text{ProfitAfterTax} &= \text{ProfitBeforeTax} - \text{ProfitTax},
\end{align*}

so that a loss year triggers no tax liability but also earns no rebate,
mirroring the asymmetric treatment a real small proprietorship faces. On the
three year horizon this ledger additionally tracks retained earnings, owner
withdrawals, one time capital expenditures, and the cash payment of the
prior year's profit tax, exactly the additional bookkeeping that the
expansion decision described in Section~\ref{sec:phase5} depends upon.

\section{Discussion}
\label{sec:discussion}

Several design choices recur across every phase of this model and are worth
drawing out explicitly, because they explain why the architecture as a whole
looks the way it does rather than some equally plausible alternative.

The first recurring choice is that every random variable is defended on its
own terms by the same three question discipline introduced in
Section~\ref{sec:principles}, rather than by convention or convenience. This
has a practical consequence beyond intellectual tidiness: because each
distributional choice is defended by an explicit generative story, the size
and shape of every planted pattern in the resulting data, such as the degree
of overstocking caused by owner optimism or the seasonal swing in a
particular product's sales, can in principle be computed exactly from the
model's own parameters rather than merely observed empirically after the
fact.

The second recurring choice is the strict separation between what is
exogenous, meaning decided in advance and independent of the simulated
agents' behavior, and what is endogenous, meaning a genuine response to
observed outcomes within the model. Weather, macroeconomic shocks, and
household budget shocks are all exogenous by construction, which is what
allows them to function as instruments and natural experiments in later
analysis. The owner's restocking rule, pricing rule, and expansion decision
are all endogenous responses to what he actually observes, which is what
allows his documented mistakes to be genuine, traceable consequences of
otherwise reasonable behavior rather than arbitrary flaws inserted for
pedagogical convenience.

The third recurring choice is the strict separation between the underlying
physical reality of the simulated world, which is always internally
consistent and exactly conserved, and the paperwork describing that reality,
which is allowed to be imperfect in the same limited, realistic ways real
business paperwork is imperfect. This separation is what allows the
counterfactual machinery of Phases Four and Five to remain exact even in the
presence of realistic recording noise, because that noise is injected only
at the point of export and never touches the simulated state itself.

Finally, throughout the model a small number of specific, deliberate gaps
between belief and truth, whether the owner's demand estimate, his
forecasting rule, his tax strategy, or his capital allocation decision, are
each introduced once, defended on realistic grounds, and then left
undisturbed rather than corrected. This is a deliberate design stance rather
than an oversight: the entire point of the construction is to produce a
world in which the distance between what an agent believes and what is true
can be measured precisely, and that distance can only be measured if the
model resists the temptation to make its own protagonist smarter than the
premise requires.

\section{Conclusion}
\label{sec:conclusion}

This paper has consolidated the theoretical reasoning behind a five phase
construction of a synthetic grocery store, from the owner's opening decision
solved as a mixed integer program, through a scripted year of weather and
macroeconomic shocks, a daily market mechanism built on individually
rational discrete choice, a laboratory of counterfactual scenarios, and a
three year extension adding customer turnover and capital accumulation. Every
distributional and structural choice throughout the model was defended by
the same discipline: identifying the variable's true support, articulating
the generative story that actually produces it, and selecting, within those
constraints, the distribution that assumes the least beyond what has
actually been claimed. The result is a world built from first principles
rather than a spreadsheet of plausible numbers, in which every observable
pattern, whether a seasonal swing in ice cream sales, an overstocked
category, or a capital decision that looks reasonable at the time and
questionable in hindsight, traces to an explicit, defensible mechanism rather
than to an arbitrary injection of noise.

\section*{References}
\addcontentsline{toc}{section}{References}

\begin{list}{}{\leftmargin=1.5em \itemindent=-1.5em \itemsep=8pt}

\item \hypertarget{ref:dwork2006}{}Dwork, C. (2006). Differential privacy.
In \textit{Proceedings of the 33rd International Colloquium on Automata,
Languages and Programming (ICALP)}, Part II, Lecture Notes in Computer
Science vol.\ 4052, 1--12. Springer.

\item \hypertarget{ref:mcfadden1974}{}McFadden, D. (1974). Conditional
logit analysis of qualitative choice behavior. In P. Zarembka (Ed.),
\textit{Frontiers in Econometrics}, 105--142. Academic Press.

\item \hypertarget{ref:orcutt1957}{}Orcutt, G. H. (1957). A new type of
socio-economic system. \textit{Review of Economics and Statistics}, 39(2),
116--123.

\item \hypertarget{ref:patki2016}{}Patki, N., Wedge, R., and
Veeramachaneni, K. (2016). The synthetic data vault. In \textit{Proceedings
of the IEEE International Conference on Data Science and Advanced Analytics
(DSAA)}, 399--410.

\item \hypertarget{ref:xu2019}{}Xu, L., Skoularidou, M., Cuesta-Infante,
A., and Veeramachaneni, K. (2019). Modeling tabular data using conditional
GAN. In \textit{Advances in Neural Information Processing Systems 32
(NeurIPS 2019)}, 7335--7345.

\end{list}

\end{document}
